---
title: "spatial-separation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spatial-separation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # Usage guide only
)
```

```{r setup}
library(climate.risk.tool)

base_dir <- file.path("..", "tests", "tests_data")
```

## Prepare inputs

```{r}
assets <- read_assets(file.path(base_dir, "user_input"))
companies <- read_companies(file.path(base_dir, "user_input"))

hazard_data <- load_hazards_and_inventory(
  hazards_dir = file.path(base_dir, "hazards", "config"),
  hazard_indicators_dir = file.path(base_dir, "hazards", "indicators"),
  aggregate_factor = 16L
)

hazards <- c(hazard_data$hazards$tif, hazard_data$hazards$nc, hazard_data$hazards$csv)
hazards_inventory <- hazard_data$inventory
hazard_configs <- hazard_data$configs
hazards_dir <- file.path(base_dir, "hazards", "config")
precomputed_hazards <- read_precomputed_hazards(base_dir)
```

## Define events (default behavior: no spatial separation)

When no spatial columns are provided, `compute_risk()` keeps the default behavior and does not spatially separate assets.

```{r}
events <- data.frame(
  event_id = c("ev1", "ev2", "ev3", "ev4"),
  hazard_type = c("Flood", "Flood", "Heat", "Heat"),
  hazard_name = c(
    "Flood__depth(cm)__GWL=CurrentClimate__RP=10",
    "Flood__depth(cm)__GWL=CurrentClimate__RP=10",
    "Compound__HI__GWL=present__RP=10__ensemble=mean",
    "Compound__HI__GWL=3__RP=10__ensemble=mean"
  ),
  scenario_name = c("CurrentClimate", "CurrentClimate", "present", "3"),
  return_period = c(10, 10, 10, 10),
  event_year = c(2030L, 2030L, 2030L, 2035L),
  stringsAsFactors = FALSE
)
```

```{r}
results_default <- compute_risk(
  assets = assets,
  companies = companies,
  events = events,
  hazards = hazards,
  hazards_inventory = hazards_inventory,
  precomputed_hazards = precomputed_hazards,
  hazard_configs = hazard_configs,
  hazards_dir = hazards_dir,
  growth_rate = 0.02,
  discount_rate = 0.05,
  risk_free_rate = 0.02
)
```

## Apply per-event spatial separation programmatically

Use `set_events_spatial_separation()` to assign filters per `event_id`.

```{r}
events_spatial <- set_events_spatial_separation(
  events = events,
  hazard_configs = hazard_configs,
  spatial_filters = list(
    # ADM scheme shorthand (state-level by codes)
    ev3 = list(state = c("11", "33")),
    # Hydro scheme shorthand (micro hydrological region codes)
    ev1 = list(micro = c("10316084", "11151279")),
    # Explicit form with labels and level
    ev4 = list(
      level = "municipality",
      codes = c("1100023", "1200401"),
      labels = c("Ariquemes", "Rio Branco")
    )
  )
)

events_spatial
```

## Run with spatial separation

```{r}
results_spatial <- compute_risk(
  assets = assets,
  companies = companies,
  events = events_spatial,
  hazards = hazards,
  hazards_inventory = hazards_inventory,
  precomputed_hazards = precomputed_hazards,
  hazard_configs = hazard_configs,
  hazards_dir = hazards_dir,
  growth_rate = 0.02,
  discount_rate = 0.05,
  risk_free_rate = 0.02
)
```

Inspect status columns added by spatial filtering:

```{r}
results_spatial$assets_factors |>
  dplyr::select(asset, event_id, hazard_type, spatial_exposure_status, spatial_multiplier)
```

